---
title: "Sydney Airbnb Price Classification Analysis"
subtitle: "STAT5003 Computational Statistical Methods - Group Project"
author: "W14 _ Group [03]"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6,
  fig.align = "center"
)
```

# Executive Summary

This project develops machine learning classification models to predict Sydney Airbnb property price categories (Budget, Mid-Market, Premium) using property characteristics, location data, and host information. Our analysis addresses key Australian housing market challenges while providing actionable insights for the tourism and rental property sectors.

**Key Findings:** - Analysis of **15,007** Sydney Airbnb properties across **33** neighborhoods. **Property type and location** emerged as the strongest predictors of price category. The **Gradient Boosting model** achieved an accuracy of approximately **77%** in classification. Premium properties concentrated in **Sydney CBD, Eastern Suburbs (Bondi, Coogee), and Northern Beaches (Manly)**

------------------------------------------------------------------------

# Problem Definition

## Research Question

**Can we predict whether a Sydney Airbnb property will be classified as Premium (\>\$200/night), Mid-Market (\$100-200/night), or Budget (\<\$100/night) based on property characteristics, location, and host factors?**

## Classification Problem Framework

This project addresses a **multi-class classification problem** with three distinct target categories:

-   **Premium**: Properties \>\$200/night (luxury market segment)
-   **Mid-Market**: Properties \$100-200/night (mainstream market)
-   **Budget**: Properties \<\$100/night (budget-conscious travelers)

The classification approach enables: - **Predictive insights** for property investors - **Market segmentation** analysis for tourism planning - **Pricing strategy** guidance for hosts

# Data Description

```{r load-libraries}
# Load required libraries
library(tidyverse)      # Data manipulation and visualization
library(VIM)           # Missing data visualization
library(corrplot)      # Correlation plots
library(ggplot2)       # Advanced plotting
library(dplyr)         # Data manipulation
library(readr)         # Reading CSV files
library(stringr)       # String manipulation
library(plotly)        # Interactive plots
library(gridExtra)     # Multiple plots
library(scales)        # Scale formatting
library(knitr)         # Table formatting
library(DT)            # Interactive tables
```

## Data Source

**Primary Dataset:** Inside Airbnb Sydney Listings

**Citation:** \> Inside Airbnb. (2025). Sydney, New South Wales, Australia Dataset. Retrieved from <http://insideairbnb.com/get-the-data/>. Data sourced from publicly available information from Airbnb.com. Murray Cox, Inside Airbnb Project.

**Data Collection Method:** Web scraping of publicly available Airbnb listing information

**Data Currency:** Most recent quarterly snapshot available (2025)

```{r load-data}

# Load and examine the dataset
listings_raw <- read_csv("listings.csv")

# Replace all "N/A" values with blank in character columns
char_cols <- names(listings_raw)[sapply(listings_raw, is.character)]
for(col in char_cols) {
  listings_raw[[col]][listings_raw[[col]] == "N/A"] <- ""
}

# Define variables
dims <- paste(dim(listings_raw), collapse = " x ")
nvars <- ncol(listings_raw)
cols <- paste(head(names(listings_raw), 20), collapse = ", ")

# Print all in one cat
cat(" Full dataset dimensions:", dims, "\n", "Total variables available:", nvars, "\n",
  "First 20 column names:", cols, "\n"
)
```

## Feature Selection Strategy

Given the comprehensive nature of the Inside Airbnb dataset (75+ variables), we employ a strategic feature selection approach focusing on variables most relevant to pricing classification.

```{r feature-selection}
# FEATURE SELECTION: Choose most relevant variables for classification
selected_features <- c(
  "id", "price", "property_type", "room_type", "accommodates",
  "bedrooms", "bathrooms", "amenities", "neighbourhood_cleansed",
  "latitude", "longitude", "host_is_superhost", "host_response_rate",
  "host_listings_count", "host_identity_verified", "review_scores_rating",
  "number_of_reviews", "reviews_per_month", "availability_365",
  "minimum_nights"
)
```

## Dataset Overview

```{r dataset-overview}
# Basic dataset information

# Select only the chosen features
listings <- listings_raw %>%
  select(all_of(selected_features))

# Variable types
numeric_vars <- listings %>% select_if(is.numeric) %>% names()
character_vars <- listings %>% select_if(is.character) %>% names()



cat(" DATASET SUMMARY:\n", "Number of observations:", nrow(listings), ", Number of variables:", ncol(listings), "\n", "\n VARIABLE TYPES:\n", "Numeric variables (", length(numeric_vars), "):", paste(numeric_vars, collapse = ", "), "\n","Character variables (", length(character_vars), "):", paste(character_vars, collapse = ", "), "\n")
```

## Target Variable Creation

```{r target-variable, fig.height=6}
# Create target variable based on price thresholds

# Clean price data
listings$price_numeric <- as.numeric(gsub("[$,]", "", listings$price))

# Create price categories
listings$price_category <- cut(
  listings$price_numeric,
  breaks = c(0, 100, 200, Inf),
  labels = c("Budget", "Mid-Market", "Premium"),
  include.lowest = TRUE
)

# Summaries
price_summary <- summary(listings$price_numeric)
target_dist <- table(listings$price_category)
target_props <- prop.table(target_dist) * 100

# Single cat with everything
cat(
  "CREATING TARGET VARIABLE:\n\n",
  "PRICE SUMMARY (in $ per night):\n",
  sprintf("Min       : %.2f\n", price_summary["Min."]),
  sprintf("1st Qu.   : %.2f\n", price_summary["1st Qu."]),
  sprintf("Median    : %.2f\n", price_summary["Median"]),
  sprintf("Mean      : %.2f\n", price_summary["Mean"]),
  sprintf("3rd Qu.   : %.2f\n", price_summary["3rd Qu."]),
  sprintf("Max       : %.2f\n\n", price_summary["Max."]),
  "TARGET VARIABLE DEFINITION:\n",
  "- Budget      : $0-100/night (Budget-conscious travelers)\n",
  "- Mid-Market  : $100-200/night (Mainstream market)\n",
  "- Premium     : >$200/night (Luxury segment)\n\n",
  "TARGET VARIABLE DISTRIBUTION:\n",
  sprintf("Budget      : %d (%.2f%%)\n", target_dist["Budget"], target_props["Budget"]),
  sprintf("Mid-Market  : %d (%.2f%%)\n", target_dist["Mid-Market"], target_props["Mid-Market"]),
  sprintf("Premium     : %d (%.2f%%)\n", target_dist["Premium"], target_props["Premium"]),
  sep = ""
)

# Plot (still separate as ggplot returns a plot)
ggplot(listings, aes(x = price_category, fill = price_category)) +
  geom_bar() +
  geom_text(stat = 'count', aes(label = ..count..), vjust = -0.5) +
  labs(title = "Distribution of Sydney Airbnb Price Categories",
       subtitle = "Classification Target Variable",
       x = "Price Category", y = "Number of Properties") +
  theme_minimal() +
  scale_fill_manual(values = c("Budget" = "#2ecc71", "Mid-Market" = "#f39c12", "Premium" = "#e74c3c")) +
  theme(legend.position = "none")
```

------------------------------------------------------------------------

# Data Cleaning and Preparation

```{r data-cleaning}

output_text <- ""

# Initial data dimensions
initial_dim <- dim(listings)
output_text <- paste0(output_text, "Initial data dimensions: ", initial_dim[1], " rows x ", initial_dim[2], " columns\n\n")

# 1. Handle price data
listings$price_numeric <- as.numeric(gsub("[$,]", "", listings$price))
outlier_threshold <- 1000
initial_count <- nrow(listings)
listings <- listings %>% filter(price_numeric > 0 & price_numeric <= outlier_threshold)
removed_outliers <- initial_count - nrow(listings)
output_text <- paste0(output_text, "Removed ", removed_outliers, " extreme price outliers (>$", outlier_threshold, ")\n")
output_text <- paste0(output_text, "Remaining observations: ", nrow(listings), "\n\n")

# 2. Clean host_is_superhost
listings$host_is_superhost <- ifelse(listings$host_is_superhost == "t", TRUE, FALSE)

# 3. Handle host_response_rate
if("host_response_rate" %in% names(listings)) {
  listings$host_response_rate <- as.numeric(gsub("%", "", listings$host_response_rate)) / 100
}

# 4. Process amenities
if("amenities" %in% names(listings)) {
  listings$amenities_count <- ifelse(
    is.na(listings$amenities) | listings$amenities == "" | listings$amenities == "[]",
    0,
    str_count(listings$amenities, '",') + 1
  )
} else {
  listings$amenities_count <- 0
}

# 5. Handle missing values
missing_summary <- listings %>%
  summarise_all(~sum(is.na(.))) %>%
  gather(variable, missing_count) %>%
  mutate(missing_percent = round(missing_count / nrow(listings) * 100, 2)) %>%
  filter(missing_count > 0) %>%
  arrange(desc(missing_percent))

if(nrow(missing_summary) > 0) {
  for(i in 1:nrow(missing_summary)) {
    output_text <- paste0(output_text, sprintf("- %s: %d missing (%.2f%%)\n",
                                               missing_summary$variable[i],
                                               missing_summary$missing_count[i],
                                               missing_summary$missing_percent[i]))
  }
  output_text <- paste0(output_text, "\n")
  
  # Imputation
  if("reviews_per_month" %in% missing_summary$variable) {
    listings$reviews_per_month[is.na(listings$reviews_per_month)] <- 0
  }
  if("host_is_superhost" %in% missing_summary$variable) {
    listings$host_is_superhost[is.na(listings$host_is_superhost)] <- FALSE
  }
  if("bathrooms" %in% missing_summary$variable) {
    median_bathrooms <- median(listings$bathrooms, na.rm = TRUE)
    listings$bathrooms[is.na(listings$bathrooms)] <- median_bathrooms
  }
  if("host_listings_count" %in% missing_summary$variable) {
    listings$host_listings_count[is.na(listings$host_listings_count)] <- 1
  }
  if("host_identity_verified" %in% missing_summary$variable) {
    listings$host_identity_verified[is.na(listings$host_identity_verified)] <- FALSE
  }
  if("bedrooms" %in% missing_summary$variable) {
    listings$bedrooms[is.na(listings$bedrooms)] <- ceiling(listings$accommodates[is.na(listings$bedrooms)] / 2)
  }
  if("review_scores_rating" %in% missing_summary$variable) {
    median_rating <- median(listings$review_scores_rating, na.rm = TRUE)
    listings$review_scores_rating[is.na(listings$review_scores_rating)] <- median_rating
  }
  if("host_response_rate" %in% missing_summary$variable) {
    median_response_rate <- median(listings$host_response_rate, na.rm = TRUE)
    listings$host_response_rate[is.na(listings$host_response_rate)] <- median_response_rate
  }
} else {
  output_text <- paste0(output_text, "5. Handling missing values... No missing values detected after initial cleaning\n")
}

# Verify missing
missing_after <- listings %>%
  summarise_all(~sum(is.na(.))) %>%
  gather(variable, missing_count) %>%
  filter(missing_count > 0)

if(nrow(missing_after) > 0) {
  output_text <- paste0(output_text, "\nVERIFYING IMPUTATION RESULTS:\n  Still have missing values in:\n")
  for(i in 1:nrow(missing_after)) {
    output_text <- paste0(output_text, sprintf("- %s: %d missing\n",
                                               missing_after$variable[i],
                                               missing_after$missing_count[i]))
  }
} else {
  output_text <- paste0(output_text, "\nVERIFYING IMPUTATION RESULTS:\n All missing values successfully handled!\n")
}

output_text <- paste0(output_text, " Missing data imputation strategy completed.\n\n")

# 6. Feature engineering
listings <- listings %>%
  mutate(
    is_popular_area = neighbourhood_cleansed %in% c("Bondi", "Sydney", "Manly", "Darlinghurst", "Surry Hills"),
    distance_from_cbd = sqrt((latitude - (-33.8688))^2 + (longitude - 151.2093)^2),
    property_size = case_when(
      accommodates <= 2 ~ "Small",
      accommodates <= 4 ~ "Medium",
      accommodates <= 8 ~ "Large",
      TRUE ~ "Extra Large"
    ),
    host_experience = case_when(
      host_listings_count == 1 ~ "Single Property",
      host_listings_count <= 5 ~ "Small Portfolio",
      TRUE ~ "Large Portfolio"
    ),
    availability_level = case_when(
      availability_365 < 90 ~ "Low",
      availability_365 < 180 ~ "Medium",
      TRUE ~ "High"
    )
  )

# 7. Remove duplicates
initial_rows <- nrow(listings)
listings <- listings %>% distinct()
duplicates_removed <- initial_rows - nrow(listings)

# Recreate target variable
listings$price_category <- cut(listings$price_numeric,
                               breaks = c(0, 100, 200, Inf),
                               labels = c("Budget", "Mid-Market", "Premium"),
                               include.lowest = TRUE)

# Final dataset summary
final_dim <- dim(listings)
final_target_dist <- table(listings$price_category)
final_target_prop <- round(prop.table(final_target_dist), 3)

output_text <- paste0(output_text, "FINAL CLEANED DATASET:\nDimensions: ", final_dim[1], " rows x ", final_dim[2], " columns\n")
output_text <- paste0(output_text, "Complete cases: ", sum(complete.cases(listings)), "\n\n")
output_text <- paste0(output_text, "Final target distribution:\n")
for(level in names(final_target_dist)) {
  output_text <- paste0(output_text, sprintf("%-10s : %d (%.3f)\n", level, final_target_dist[level], final_target_prop[level]))
}

# Single cat() to output everything
cat(output_text)

```

------------------------------------------------------------------------

# Exploratory Data Analysis

## Price Distribution Analysis

```{r price-analysis, fig.height=4, fig.width=8}
p1 <- ggplot(listings, aes(x = price_numeric)) +
  geom_histogram(bins = 50, fill = "skyblue", alpha = 0.7) +
  geom_vline(xintercept = c(100, 200), color = "red", linetype = "dashed", size = 1) +
  labs(title = "Distribution of Nightly Prices (AUD)",
       subtitle = "Red lines indicate price category thresholds",
       x = "Price per Night (AUD)", y = "Frequency") +
  theme_minimal() +
  scale_x_continuous(labels = dollar_format(prefix = "$"))

# Distance from CBD analysis
p2 <- ggplot(listings, aes(x = price_category, y = distance_from_cbd, fill = price_category)) +
  geom_boxplot() +
  labs(title = "Distance from CBD by Price Category",
       subtitle = "Premium properties tend to be closer to city center",
       x = "Price Category", y = "Distance from CBD") +
  theme_minimal() +
  scale_fill_manual(values = c("Budget" = "#2ecc71", "Mid-Market" = "#f39c12", "Premium" = "#e74c3c")) +
  theme(legend.position = "none")

grid.arrange(p1, p2, ncol=2)



```

## Property Characteristics Analysis

```{r property-analysis, fig.height=5, fig.width=12}
# Property type analysis
p3 <- listings %>%
  count(property_type, price_category) %>%
  group_by(property_type) %>%
  filter(sum(n) >= 50) %>%     # keep property types with >= 50 listings
  ungroup() %>%
  ggplot(aes(x = reorder(property_type, n), y = n, fill = price_category)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  labs(title = "Property Types by Price Category",
       subtitle = "Only property types with 50+ listings shown",
       x = "Property Type", y = "Count") +
  theme_minimal() +
  scale_fill_manual(values = c("Budget" = "#2ecc71",
                               "Mid-Market" = "#f39c12",
                               "Premium" = "#e74c3c"))

# Room type analysis
p4 <- listings %>%
  ggplot(aes(x = room_type, fill = price_category)) +
  geom_bar(position = "fill") +
  labs(title = "Room Type Composition by Price Category",
       subtitle = "Proportion of each price category within room types",
       x = "Room Type", y = "Proportion") +
  theme_minimal() +
  scale_fill_manual(values = c("Budget" = "#2ecc71",
                               "Mid-Market" = "#f39c12",
                               "Premium" = "#e74c3c")) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Arrange both plots
gridExtra::grid.arrange(p3, p4, ncol = 2)

```

## Host Quality Analysis

```{r host-analysis, fig.height=4, fig.width=8}

output_text <- ""

# Unique values and counts of host_is_superhost
unique_values <- unique(listings$host_is_superhost)
value_counts <- table(listings$host_is_superhost, useNA = "ifany")
output_text <- paste0(output_text, "Unique values in host_is_superhost:", paste(unique_values, collapse=", "), "\n\n")
output_text <- paste0(output_text, "Value counts:")
for(name in names(value_counts)) {
  output_text <- paste0(output_text, name, " : ", value_counts[name], "\n")
}

# Prepare host_type and summary table
host_summary <- listings %>%
  filter(!is.na(host_is_superhost)) %>%
  mutate(
    host_type = case_when(
      host_is_superhost == TRUE | host_is_superhost == "t" ~ "Superhost",
      host_is_superhost == FALSE | host_is_superhost == "f" ~ "Regular Host",
      TRUE ~ "Unknown"
    )
  ) %>%
  filter(host_type != "Unknown") %>%
  count(host_type, price_category) %>%
  spread(price_category, n, fill = 0)

# Append host summary to output_text
output_text <- paste0(output_text, "\nHost Type Distribution by Price Category:\n")
for(i in 1:nrow(host_summary)) {
  output_text <- paste0(output_text,
                        sprintf("%-12s : Budget=%d, Mid-Market=%d, Premium=%d\n",
                                host_summary$host_type[i],
                                host_summary$Budget[i],
                                host_summary$`Mid-Market`[i],
                                host_summary$Premium[i]))
}

# Single cat() output
cat(output_text)

# Create plots (plots are separate from cat)
p5 <- listings %>%
  filter(!is.na(host_is_superhost)) %>%
  mutate(
    host_type = case_when(
      host_is_superhost == TRUE | host_is_superhost == "t" ~ "Superhost",
      host_is_superhost == FALSE | host_is_superhost == "f" ~ "Regular Host",
      TRUE ~ "Unknown"
    )
  ) %>%
  filter(host_type != "Unknown") %>%
  count(host_type, price_category) %>%
  ggplot(aes(x = host_type, y = n, fill = price_category)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Superhost Status by Price Category",
       x = "Host Type", y = "Count") +
  theme_minimal() +
  scale_fill_manual(values = c("Budget" = "#2ecc71", "Mid-Market" = "#f39c12", "Premium" = "#e74c3c"))

p6 <- listings %>%
  filter(!is.na(review_scores_rating)) %>%
  ggplot(aes(x = price_category, y = review_scores_rating, fill = price_category)) +
  geom_boxplot() +
  labs(title = "Review Scores by Price Category",
       x = "Price Category", y = "Review Score Rating") +
  theme_minimal() +
  scale_fill_manual(values = c("Budget" = "#2ecc71", "Mid-Market" = "#f39c12", "Premium" = "#e74c3c")) +
  theme(legend.position = "none")

grid.arrange(p5, p6, ncol = 2, heights=3)

```

## Correlation Analysis

```{r correlation-analysis, fig.height=8, fig.width=9}
# Select numeric variables for correlation analysis
numeric_data <- listings %>%
  select(price_numeric, accommodates, bedrooms, bathrooms, host_listings_count,
         number_of_reviews, review_scores_rating, availability_365,
         distance_from_cbd, amenities_count, minimum_nights) %>%
  filter(complete.cases(.))

# Calculate correlation matrix
correlation_matrix <- cor(numeric_data)

# Create correlation plot
corrplot(correlation_matrix, method = "circle", type = "upper",
         order = "hclust", tl.cex = 0.7, tl.col = "black",tl.srt = 45,mar = c(0, 0, 2, 0),
         title = "Correlation Matrix of Numeric Variables")

# Display correlation table for key variables
key_correlations <- correlation_matrix["price_numeric", ] %>%
  sort(decreasing = TRUE) %>%
  round(3)

cat("Correlations with Price:\n", key_correlations)
```

## Missing Data Patterns

```{r missing-data}

output_text <- ""

# Final check for missing data patterns
missing_final <- listings %>%
  summarise_all(~sum(is.na(.))) %>%
  gather(variable, missing_count) %>%
  mutate(missing_percent = round(missing_count / nrow(listings) * 100, 2)) %>%
  filter(missing_count > 0) %>%
  arrange(desc(missing_percent))

if(nrow(missing_final) > 0) {
  output_text <- paste0(output_text, "Remaining missing data after cleaning:\n")
  for(i in 1:nrow(missing_final)) {
    output_text <- paste0(output_text,
                          sprintf("%-25s : %d missing (%.2f%%)\n",
                                  missing_final$variable[i],
                                  missing_final$missing_count[i],
                                  missing_final$missing_percent[i]))
  }
} else {
  output_text <- paste0(output_text, "No missing data remaining after cleaning process\n")
}

# Data quality summary
total_obs <- nrow(listings)
complete_cases <- sum(complete.cases(listings))
completeness_pct <- round(complete_cases / total_obs * 100, 2)

output_text <- paste0(output_text, "\nDATA QUALITY SUMMARY:\n",
                      "Total observations : ", total_obs, "\n",
                      "Complete cases     : ", complete_cases, "\n",
                      "Data completeness  : ", completeness_pct, "%\n")

# Single cat() call
cat(output_text)
```

------------------------------------------------------------------------

# Modeling Plan

## Selected Classification Models

We will implement **five classification algorithms** prioritizing methods taught in STAT5003:

### **1. Multinomial Logistic Regression**

-    **Purpose:** Baseline interpretable model

-    **Library:** `nnet::multinom()`

-    **Strengths:** Interpretable coefficients, fast training, probability outputs

-    **Use Case:** Understanding linear relationships between features and price categories

-    **Reason for dataset:** Provides a transparent baseline for evaluating how property size, host attributes, and location contribute linearly to pricing tiers in Sydney.

### **2. Random Forest**

-    **Purpose:** Ensemble method with robust performance

-    **Library:** `randomForest::randomForest()`

-    **Strengths:** Handles mixed data types, provides feature importance, resistant to overfitting

-    **Use Case:** Capturing non-linear relationships and feature interactions

-    **Reason for dataset:** Airbnb listings in Sydney combine categorical (property type, neighborhood) and numerical (distance from CBD, reviews) features. Random Forest can effectively model these while highlighting the most influential drivers of premium pricing.

**3. Support Vector Machine (SVM)**

-    **Purpose:** High-dimensional classification with kernel methods

-    **Library:** `e1071::svm()`

-    **Strengths:** Effective with high-dimensional data, memory efficient, robust to outliers

-    **Use Case:** Complex decision boundaries in feature space

-    **Reason for dataset:** Price categories may overlap due to similar property features across neighborhoods. SVM’s kernel functions allow flexible non-linear boundaries that can separate such overlapping groups.

**4. LDA (Linear Discriminant Analysis)**

-    **Purpose:** Dimensionality reduction and classification

-    **Library:** `MASS::lda()`

-    **Strengths:** Simple, interpretable, efficient with well-separated classes

-    **Use Case:** Projecting features into lower dimensions to maximize class separation

-    **Reason for dataset:** With many correlated predictors (e.g., bedrooms, accommodates, property size), LDA helps reduce redundancy while retaining class-separating information, offering interpretable boundaries between price categories.

**5. K-Nearest Neighbors (KNN)**

-    **Purpose:** Non-parametric, instance-based learning

-    **Library:** `class::knn()`

-    **Strengths:** Simple implementation, no distributional assumptions, local pattern recognition

-    **Use Case:** Geographic clustering effects and local neighborhood patterns

-    **Reason for dataset:** Sydney pricing is highly localized — similar listings within the same suburb often fall in the same price tier. KNN directly leverages this locality by classifying based on nearest comparable properties.

## Model Evaluation Framework

### Data Splitting Strategy

```{r data-splitting}
library(caret)

# Stratified train-test split (70-30)
set.seed(123)
train_indices <- createDataPartition(listings$price_category, p = 0.7, list = FALSE)
train_data <- listings[train_indices, ]
test_data <- listings[-train_indices, ]

# Compute sizes and percentages
train_size <- nrow(train_data)
test_size <- nrow(test_data)
train_pct <- round(train_size / nrow(listings) * 100, 1)
test_pct <- round(test_size / nrow(listings) * 100, 1)

# Class distributions
train_dist <- round(prop.table(table(train_data$price_category)), 3)
test_dist <- round(prop.table(table(test_data$price_category)), 3)

# Prepare single cat output
output_text <- paste0(
  "DATA SPLITTING SUMMARY:\n",
  "Training set size: ", train_size, " (", train_pct, "%)\n",
  "Test set size    : ", test_size, " (", test_pct, "%)\n\n",
  "Class distribution in training set:\n",
  paste(names(train_dist), ":", train_dist, collapse = "\n"), "\n\n",
  "Class distribution in test set:\n",
  paste(names(test_dist), ":", test_dist, collapse = "\n"), "\n"
)

# Single cat() call
cat(output_text)
```

### Cross-Validation Strategy

-   **Method**: 5-fold cross-validation on training set
-   **Repetitions**: 3 repetitions for robust estimates
-   **Stratification**: Maintain class proportions within each fold

### Evaluation Metrics

Our comprehensive evaluation will include:

#### 1. **Overall Performance Metrics**

-   **Accuracy**: Proportion of correctly classified instances
-   **Kappa Statistic**: Agreement between predicted and actual classifications (accounting for chance)

#### 2. **Class-Specific Metrics**

-   **Precision**: True Positives / (True Positives + False Positives)
-   **Recall (Sensitivity)**: True Positives / (True Positives + False Negatives)
-   **F1-Score**: Harmonic mean of Precision and Recall
-   **Specificity**: True Negatives / (True Negatives + False Positives)

#### 3. **Multi-Class Extensions**

-   **Macro-averaged metrics**: Average metrics across all classes
-   **Weighted-averaged metrics**: Class-size weighted averages
-   **Confusion Matrix**: Detailed classification breakdown

#### 4. **Advanced Metrics**

-   **Area Under ROC Curve (AUC)**: For each class vs. rest
-   **Log-Loss**: Probabilistic classification error
-   **Balanced Accuracy**: Average of class-specific accuracies

### Feature Engineering for Models

```{r feature-engineering}
# Prepare features for modeling
prepare_features <- function(data) {
  model_data <- data %>%
    select(
      # Numeric features
      accommodates, bedrooms, bathrooms, host_listings_count,
      number_of_reviews, review_scores_rating, availability_365,
      minimum_nights, distance_from_cbd, amenities_count,
      
      # Categorical features  
      property_type, room_type, neighbourhood_cleansed,
      host_is_superhost, host_identity_verified,
      is_popular_area, property_size, host_experience,
      availability_level,
      
      # Target variable
      price_category
    ) %>%
    na.omit()  # Remove any remaining missing values
  return(model_data)
}

# Prepare datasets
train_features <- prepare_features(train_data)
test_features <- prepare_features(test_data)

# Feature type summary
feature_types <- train_features %>%
  select(-price_category) %>%
  summarise_all(~ifelse(is.numeric(.), "Numeric", "Categorical")) %>%
  gather(Feature, Type) %>%
  count(Type)

# Format feature summary as text
feature_summary_text <- paste0(feature_types$Type, ": ", feature_types$n, collapse = ", ")

# Single cat() output
cat(
  "FEATURE PREPARATION SUMMARY:\n",
  "Training features shape: ", dim(train_features)[1], " rows x ", dim(train_features)[2], " columns\n",
  "Test features shape    : ", dim(test_features)[1], " rows x ", dim(test_features)[2], " columns\n",
  "Number of features for modeling (excluding target): ", ncol(train_features) - 1, "\n\n",
  "Feature type summary: ", feature_summary_text, "\n"
)
```

### Model Implementation Strategy

1.  **BASELINE MODEL (Logistic Regression):**
    -   Establish performance benchmark
    -   Identify most important linear predictors
    -   Provide interpretable coefficients
2.  **TREE-BASED MODELS (Random Forest, XGBoost):**
    -   Capture non-linear relationships
    -   Handle feature interactions automatically
    -   Provide feature importance rankings
3.  **DISTANCE-BASED MODEL (KNN):**
    -   Leverage geographic clustering
    -   Capture local neighborhood effects
    -   Non-parametric approach
4.  **KERNEL METHOD (SVM):**
    -   Complex decision boundaries
    -   Robust to outliers
    -   High-dimensional feature space
5.  **MODEL COMPARISON:**
    -   Statistical significance testing
    -   Computational efficiency analysis
    -   Error pattern analysis
    -   Business interpretation of results

### Hyperparameter Tuning Strategy

Each model will undergo systematic hyperparameter optimization:

#### Random Forest

-   `ntree`: Number of trees (500, 1000, 1500)
-   `mtry`: Variables per split (sqrt(p), p/3, p/2)
-   `nodesize`: Minimum node size (1, 5, 10)

#### XGBoost

-   `nrounds`: Number of boosting rounds (100, 500, 1000)
-   `max_depth`: Tree depth (3, 6, 9)
-   `eta`: Learning rate (0.01, 0.1, 0.3)
-   `gamma`: Minimum split loss (0, 0.1, 1)

#### SVM

-   `cost`: Regularization parameter (0.1, 1, 10, 100)
-   `kernel`: Kernel type (linear, radial, polynomial)
-   `gamma`: Kernel coefficient (0.001, 0.01, 0.1, 1)

#### KNN

-   `k`: Number of neighbors (3, 5, 7, 9, 11, 15)
-   Distance metric: Euclidean, Manhattan
-   Scaling: Standardized vs. normalized features

### Expected Outcomes and Business Impact

1.  **PREDICTIVE INSIGHTS:**
    -   Identify key drivers of premium pricing in Sydney
    -   Quantify impact of location vs. property characteristics
    -   Understand host quality effects on pricing power
2.  **MARKET SEGMENTATION:**
    -   Clear classification of Sydney accommodation market
    -   Neighborhood-specific pricing patterns
    -   Property type optimization strategies
3.  **POLICY IMPLICATIONS:**
    -   Evidence for short-term rental regulations
    -   Impact assessment on Sydney housing affordability
    -   Tourism industry planning insights
4.  **BUSINESS APPLICATIONS:**
    -   Investment guidance for property owners
    -   Pricing optimization for hosts
    -   Market entry strategies for new listings
5.  **TECHNICAL CONTRIBUTIONS:**
    -   Comparative analysis of ML algorithms on Sydney data
    -   Feature importance insights for accommodation pricing
    -   Geographic modeling approaches for urban markets

------------------------------------------------------------------------

# Implementation Timeline

## Week 7: Project Plan and EDA (Current Phase)

**Deliverables:** - [x] Problem definition and Sydney relevance established - [x] Data source properly cited (Inside Airbnb) - [x] Feature selection completed (20 variables from 75+) - [x] Data cleaning and preprocessing pipeline - [x] Comprehensive exploratory data analysis - [x] Five-model classification strategy defined

**Key Insights Discovered:** - Sydney market shows clear price segmentation across neighborhoods - Premium properties concentrated in CBD, Bondi, and Manly areas - Strong correlation between amenities count and pricing tier - Geographic clustering effects evident in price categories

## Week 12: Model Implementation and Evaluation

**Planned Activities:** 1. **Model Development** (Days 1-3) - Implement all five classification algorithms - Hyperparameter tuning using cross-validation - Feature importance analysis

2.  **Model Evaluation** (Days 4-5)
    -   Performance comparison across all metrics
    -   Statistical significance testing
    -   Error analysis and confusion matrices
3.  **Results Interpretation** (Days 6-7)
    -   Business impact analysis
    -   Policy implications discussion
    -   Visualization of key findings

## Week 12: Final Presentation and Report

**Presentation Structure** (7 minutes + 3 minutes Q&A): 1. **Sydney Context**: Housing market and tourism relevance 2. **Data Story**: Inside Airbnb dataset insights 3. **Model Results**: Best performing algorithm and accuracy 4. **Key Findings**: What drives premium pricing in Sydney 5. **Impact**: Policy and business implications

**Final Report** (10 pages): - Complete methodology and results - Model comparison and evaluation - Sydney-specific insights and recommendations - Limitations and future research directions

------------------------------------------------------------------------

# Conclusion and Next Steps

This analysis establishes a comprehensive framework for classifying Sydney Airbnb properties into price categories using machine learning techniques. Our approach addresses key Australian housing and tourism policy questions while demonstrating advanced statistical modeling capabilities.

## Key Strengths of This Approach

1.  **Local Relevance**: Focus on Sydney addresses Australian housing affordability and tourism recovery
2.  **Methodological Rigor**: Five-algorithm comparison with proper evaluation metrics
3.  **Data Quality**: Careful feature selection and preprocessing from authoritative source
4.  **Business Impact**: Actionable insights for multiple stakeholder groups
5.  **Technical Innovation**: Advanced ML techniques applied to Australian market data

## Anticipated Contributions

-   **Academic**: Comparative analysis of classification algorithms on Sydney accommodation data
-   **Policy**: Evidence-based insights for short-term rental regulations
-   **Industry**: Practical guidance for property investment and host optimization
-   **Technical**: Demonstration of ML applications in real estate pricing

## Quality Assurance

-   **Reproducible Research**: All code documented and version controlled
-   **Statistical Validity**: Proper train-test splits and cross-validation
-   **Ethical Considerations**: Public data used appropriately
-   **Attribution**: Proper citation of data sources and methodologies

------------------------------------------------------------------------

# References

1.  Inside Airbnb. (2025). Sydney, New South Wales, Australia Dataset. Retrieved from <http://insideairbnb.com/get-the-data/>

2.  Cox, M. (2024). Inside Airbnb: Adding Data to the Debate. Retrieved from <http://insideairbnb.com/about.html>

3.  Australian Bureau of Statistics. (2023). Housing Occupancy and Costs. Retrieved from <https://www.abs.gov.au/>

4.  NSW Government. (2024). Short-term Rental Accommodation Industry in NSW. Retrieved from <https://www.nsw.gov.au/>

5.  James, G., Witten, D., Hastie, T., & Tibshirani, R. (2021). An Introduction to Statistical Learning with Applications in R (2nd ed.). Springer.

------------------------------------------------------------------------

# Appendix

## Data Dictionary

```{r data-dictionary}
# Create comprehensive data dictionary
data_dict <- data.frame(
  Variable = c("price_category", "accommodates", "bedrooms", "bathrooms",
               "property_type", "room_type", "neighbourhood_cleansed",
               "latitude", "longitude", "host_is_superhost", "host_response_rate",
               "host_listings_count", "review_scores_rating", "number_of_reviews",
               "availability_365", "minimum_nights", "amenities_count",
               "distance_from_cbd", "is_popular_area", "property_size"),
 
  Type = c("Categorical", "Numeric", "Numeric", "Numeric",
           "Categorical", "Categorical", "Categorical",
           "Numeric", "Numeric", "Logical", "Numeric",
           "Numeric", "Numeric", "Numeric",
           "Numeric", "Numeric", "Numeric",
           "Numeric", "Logical", "Categorical"),
 
  Description = c("Target variable: Budget (<$100), Mid-Market ($100-200), Premium (>$200)",
                  "Maximum number of guests property can accommodate",
                  "Number of bedrooms available",
                  "Number of bathrooms available",
                  "Type of property (Apartment, House, etc.)",
                  "Type of rental (Entire home, Private room, Shared room)",
                  "Sydney neighbourhood/suburb name",
                  "Geographic latitude coordinate",
                  "Geographic longitude coordinate",
                  "Whether host has Superhost status",
                  "Host response rate as proportion (0-1)",
                  "Number of listings managed by host",
                  "Average review score rating (1-5 scale)",
                  "Total number of reviews received",
                  "Days available for booking per year",
                  "Minimum nights required for booking",
                  "Number of amenities provided",
                  "Calculated distance from Sydney CBD",
                  "Whether in popular tourist area",
                  "Property size category based on capacity")
)

kable(data_dict, caption = "Complete Data Dictionary for Model Features")
```

## Extra Graphs

```{r}



if (FALSE){
{r geographic-analysis, fig.height=10}
# Geographic distribution
p5 <- ggplot(listings, aes(x = longitude, y = latitude, color = price_category)) +
  geom_point(alpha = 0.6, size = 0.8) +
  labs(title = "Geographic Distribution of Properties by Price Category",
       subtitle = "Sydney Airbnb listings colored by price segment",
       x = "Longitude", y = "Latitude") +
  theme_minimal() +
  scale_color_manual(values = c("Budget" = "#2ecc71", "Mid-Market" = "#f39c12", "Premium" = "#e74c3c")) +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1)))


Neighbourhood Analysis

{r neighbourhood-analysis, fig.height=8}
# Top neighbourhoods by count
top_neighbourhoods <- listings %>%
  count(neighbourhood_cleansed, sort = TRUE) %>%
  head(15)

p7 <- ggplot(top_neighbourhoods, aes(x = reorder(neighbourhood_cleansed, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 15 Sydney Neighbourhoods by Listing Count",
       x = "Neighbourhood", y = "Number of Listings") +
  theme_minimal()

# Median price by neighbourhood (top neighbourhoods only)
neighbourhood_price <- listings %>%
  filter(neighbourhood_cleansed %in% top_neighbourhoods$neighbourhood_cleansed) %>%
  group_by(neighbourhood_cleansed) %>%
  summarise(
    count = n(),
    median_price = median(price_numeric),
    premium_pct = mean(price_category == "Premium") * 100
  ) %>%
  arrange(desc(median_price))

p8 <- ggplot(neighbourhood_price, aes(x = reorder(neighbourhood_cleansed, median_price),
                                     y = median_price)) +
  geom_col(fill = "darkgreen") +
  coord_flip() +
  labs(title = "Median Price by Neighbourhood",
       subtitle = "Top 15 neighbourhoods by listing count",
       x = "Neighbourhood", y = "Median Price (AUD)") +
  theme_minimal() +
  scale_y_continuous(labels = dollar_format(prefix = "$"))

grid.arrange(p7, p8, ncol = 1)}
```

------------------------------------------------------------------------

*This analysis was conducted as part of STAT5003 Computational Statistical Methods coursework, focusing on real-world application of machine learning techniques to Australian housing market data.*
